Hamiltonian mechanics is conservation of information entropy

### 2019-06-12 Articles/HamiltonianMechanicsIsConservationOfEntropy/article.tex
Add discussion on point particles

>>>>

\subsection*{Point particles vs infinitesimal parts}

At this point one may still argue that point particles and infinitesimal parts are both abstract concepts, both factually incorrect (i.e. planets are neither points nor made of infinitesimal parts), so why should we prefer one concept over the other?

Proper understanding of a physical theory means answering one question: why does the theory work? And the answer cannot be: because it does experimentally. The problem is not that the answer begs the question; it is that no theory works experimentally for all possible systems in all possible settings. In this sense, there are no ``right" theories, only ones that agree with experiments in a set of circumstances. The question, then, should be really taken to mean: what are the conditions under which the theory works? What are the assumptions or abstractions implicit in the model and where do they apply?

To us, this is not an abstract intellectual question. It is a practical matter: I have a system to study and have a choice of different models (e.g. Hamiltonian mechanics, Newtonian mechanics, quantum mechanics, Markov processes, ...), how do I know which one to pick?

In this light, if Hamiltonian mechanics is characterized as describing point-like objects, whose state is described by position and momentum, and that move according to a specified set of equations, it is not clear why it would apply to a planet and to a speck of dust in a vacuum, but not to a speck of dust in a fluid or to an electron. The size does not seem to matter and it is not clear why those equations are important: why should a point-particle be constrained to follow those equations?

If we instead characterize Hamiltonian mechanics as the deterministic and reversible motion of objects reducible to infinitesimal parts, we understand when the model choice is appropriate. Can we pretend that the system we are studying, under the conditions we are studying it, is made of infinitesimal parts that can be studied independently? In the case of an electron the answer is no. In the case of a planet or a speck of dust, if we are studying only the overall motion, then yes. Can we pretend that the evolution is deterministic and reversible? In the case of the speck of dust in a fluid, no. For a planet and speck of dust in a vacuum, yes.

The characterization of classical mechanics as describing point-particles, then, fails spectacularly at answering the most basic question: what is a classical system? Because knowing what a classical system is means knowing when classical mechanics is applicable. And saying that classical mechanics describes point-particles does not clarify when classical mechanics is applicable. Our characterization, instead, is very straightforward: a classical system is one that can be considered to be infinitesimally reducible. It is furthermore Hamiltonian if the evolution is deterministic and reversible.\footnote{In the same vein, \cite{AoPPhy1} finds that it is Newtonian if it is purely kinematic (i.e. spatial motion identify the state), it is Lagrangian if it is both Hamiltonian and Newtonian and so on.}

Also note that the approach presented leads to both the structure of phase space and the equations of motion. There is no freedom to choose other structures. With point-particles, instead, it is not clear why, for example, position is not sufficient to specify the state, or why trajectories that do not obey Hamilton's equations must be discarded. Therefore the point-particle characterization has far less explanatory power.
